### Niche detection using BANKSY

```{r}
### combine all samples

sp_list <- list()
for (i in 1:length(spe_files)) {
    sp_list[i] <- readRDS(paste0('/dski/nobackup/beileib/GHIST2/prediction2_spe/', spe_files[i]))
}

spe_joint <- do.call(cbind, sp_list)
saveRDS(spe_joint, 'spe_joint.rds')

# Run BANKSY

set.seed(123)

sample_ids <- unique(spe_joint$sample_id)
spe_bank <- lapply(sample_ids, function(id) {
  spe_subset <- spe_joint[, spe_joint$sample_id == id]
  Banksy::computeBanksy(
    spe_subset,
    assay_name = "logcounts",
    k_geom = 18
  )
})

spe_joint_bank <- do.call(cbind, spe_bank)

spe_joint_bank <- Banksy::runBanksyPCA(
  spe_joint_bank,
  lambda = 0.8,
  npcs = 30,
  group = "sample_id"
)

### 10 clusters

spe_joint_bank <- Banksy::clusterBanksy(
  spe_joint_bank,
  lambda = 0.8,
  npcs = 30,
  algo = "kmeans",
  kmeans.centers = 10
)

clust_col <- grep("^clust_", colnames(colData(spe_joint_bank)), value = TRUE)[1]
spe_joint$banksy_region <- colData(spe_joint_bank)[, clust_col]
```

### Generating Figure 5c

```{r}
celltype_cols <- c(
  "Epithelial"   = "#D1495B",
  "Fibroblast"   = "#EDAE49",
  "Perivascular" = "#A8A44A",
  "Endothelial"  = "#6A994E",
  "Macrophage"   = "#72B7A1",
  "Dendritic"    = "#4C8DB8",
  "T cell"       = "#5F7FA3",
  "NK/ILC"       = "#8E6C88",
  "B cell"       = "#B784A7",
  "Plasma"       = "#C96D8C"
)

ggplot(
  df_prop_niche,
  aes(
    x = factor(banksy_region_10),
    y = prop,
    fill = cell_type
  )
) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = celltype_cols) +
  scale_y_continuous(
    limits = c(0, 1),
    expand = c(0, 0),
    labels = scales::label_percent(accuracy = 1)
  ) +
  labs(
    x = "Niche (BANKSY region)",
    y = "Cell type proportion",
    fill = "Cell type"
  ) +
  theme_bw() +
  theme(
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.5
    ),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    
    # Axis labels + tick labels bold
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    
    legend.position = "top"
  ) +
  guides(
    fill = guide_legend(nrow = 2, byrow = TRUE)
  )
```


### Generating Figure 5d

```{r}
###############
## Figure 5d ##
###############

### test how T cell and macrophage proportion change across age
df_prop_niche <- metadata %>%
  group_by(sample_id, age, banksy_region, scClassify) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(sample_id, banksy_region) %>%
  mutate(prop = n / sum(n))

df_tcell <- df_prop_niche %>%
  filter(
    banksy_region_10 == 3,
    cell_type == "T cell"
  ) %>%
  mutate(niche = "Niche 3")

p_tcell <- ggplot(
  df_tcell,
  aes(x = age, y = prop)
) +
  geom_point(
    size = 2.5,
    alpha = 0.7,
    colour = "#4C8DB8"
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    colour = "#4C8DB8",
    fill = "#4C8DB8",
    alpha = 0.20,
    linewidth = 1
  ) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = "beta = -6.02e-04\np = 0.0042",
    hjust = 1.1,
    vjust = 1.3,
    fontface = "bold",
    size = 4
  ) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.12))
  ) +
  labs(
    x = "Age",
    y = "T cell proportion"
  ) +
  facet_wrap(~ niche) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    
    # facet 标题
    strip.background = element_rect(
      fill = "white",
      colour = "black",
      linewidth = 0.8
    ),
    strip.text = element_text(
      face = "bold",
      size = 13
    ),
    
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.5
    )
  )

p_tcell

df_macrophage <- df_prop_niche %>%
  filter(
    banksy_region_10 == 4,
    cell_type == "Macrophage"
  ) %>%
  mutate(niche = "Niche 4")

p_macrophage <- ggplot(
  df_macrophage,
  aes(x = age, y = prop)
) +
  geom_point(
    size = 2.5,
    alpha = 0.7,
    colour = "#6A994E"
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    colour = "#6A994E",
    fill = "#6A994E",
    alpha = 0.20,
    linewidth = 1
  ) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = "beta = 1.48e-03\np = 0.0257",
    hjust = 1.1,
    vjust = 1.3,
    fontface = "bold",
    size = 4
  ) +
  scale_y_continuous(
    labels = scales::label_percent(accuracy = 0.1),
    expand = expansion(mult = c(0.05, 0.12))
  ) +
  labs(
    x = "Age",
    y = "Macrophage proportion"
  ) +
  facet_wrap(~ niche) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    
    strip.background = element_rect(
      fill = "white",
      colour = "black",
      linewidth = 0.8
    ),
    strip.text = element_text(
      face = "bold",
      size = 13
    ),
    
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.5
    )
  )

p_macrophage

library(patchwork)

p_combined <- p_tcell + p_macrophage +
  plot_layout(ncol = 2)

p_combined

```


### Figure 5f: ECM remodelling in microniches which were defined in epithelial compartment

```{r}
### assign cells to supervised learning compartment (epithelial)
### nbt features
ld <- list.dirs('/dski/nobackup/beileib/GHIST2/nbt_features/', full.names = F)
ld <- ld[-1]

ld1 <- sapply(
  strsplit(ld, "-"),
  function(x) paste(x[1:2], collapse = "-")
)

ld <- ld[match(sampleid, ld1)]
ld1 <- ld1[match(sampleid, ld1)]


spe_joint$x = spe_joint@int_colData@listData[["spatialCoords"]][, 1]
spe_joint$y = spe_joint@int_colData@listData[["spatialCoords"]][, 2]

cell_df <- data.frame(
  cell_id = colnames(spe_joint),
  x = colData(spe_joint)$x,
  y = colData(spe_joint)$y,
  sample_id = colData(spe_joint)$sample_id
)

cell_df$epi_compartment <- FALSE

is_epi_patch <- function(x, y, patch_df, patch_size = 259) {

  any(
    x >= patch_df$orig_x &
    x <  patch_df$orig_x + patch_size &
    y >= patch_df$orig_y &
    y <  patch_df$orig_y + patch_size
  )
}

patch_size <- 259


for (i in 1:77) {
  patch <- read.csv(paste0('/dski/nobackup/beileib/GHIST2/nbt_features/', ld[i], '/', ld[i], '_TC_512_patch_all.csv'))
  epi_patch <- patch %>% filter(cls == 1)
  cell_sub <- which(cell_df$sample_id == ld1[i])
  hit <- mapply(function(cx, cy){

    any(
      cx >= epi_patch$orig_x &
      cx <  epi_patch$orig_x + patch_size &
      cy >= epi_patch$orig_y &
      cy <  epi_patch$orig_y + patch_size
    )

  },
  cell_df$x[cell_sub],
  cell_df$y[cell_sub]
  )
  cell_df$epi_compartment[cell_sub] <- hit
}

cell_df$epithelial_compartment <- mapply(
  is_epi_patch,
  cell_df$x,
  cell_df$y,
  MoreArgs = list(
    patch_df = epi_patch,
    patch_size = patch_size
  )
)

colData(spe_joint)$epi_compartment <- cell_df$epi_compartment
```


```{r}
### subset to epithelial compartment

spe_epi <- spe_joint[, colData(spe_joint)$epi_compartment]

### do BANKSY again with 6 clusters

set.seed(123)

sample_ids <- unique(spe_epi$sample_id)
cell_counts <- table(spe_epi$sample_id)
keep_samples <- names(cell_counts[cell_counts >= 100])

spe_epi_filtered <- spe_epi[, spe_epi$sample_id %in% keep_samples]

spe_bank <- lapply(sample_ids, function(id) {
  spe_subset <- spe_epi[, spe_epi$sample_id == id]
  Banksy::computeBanksy(
    spe_subset,
    assay_name = "logcounts",
    k_geom = 10
  )
})

spe_joint_bank <- do.call(cbind, spe_bank)

spe_joint_bank <- Banksy::runBanksyPCA(
    spe_joint_bank,
    lambda = 0.8,
    npcs = 30,
    group = "sample_id"
  )

spe_joint_bank <- Banksy::clusterBanksy(
    spe_joint_bank,
    lambda = 0.8,
    npcs = 30,
    algo = "kmeans",
    kmeans.centers = 6
  )

saveRDS(spe_joint_bank, 'spe_joint_epicompart.rds')
```


```{r}
spe_joint_epicompart <- readRDS("/dski/nobackup/beileib/GHIST2/prediction2_spe_immune/spe_joint_epicompart.rds")

df <- data.frame(
  sample_id = colData(spe_joint_epicompart)$sample_id,
  age = colData(spe_joint_epicompart)$age,
  niche = colData(spe_joint_epicompart)$clust_M0_lam0.8_kmeans6,
  ecm_score = colData(spe_joint_epicompart)$ecm_score
)

ecm_summary <- df %>%
  group_by(sample_id, age, niche) %>%
  summarise(
    mean_ecm = mean(ecm_score, na.rm = TRUE),
    .groups = "drop"
  )

reg_results <- ecm_summary %>%
  group_by(niche) %>%
  do({
    fit <- lm(mean_ecm ~ age, data = .)

    data.frame(
      beta = coef(fit)["age"],
      p_value = summary(fit)$coefficients["age", "Pr(>|t|)"]
    )
  })

reg_results <- reg_results %>%
  mutate(
    label = paste0(
      "β = ", signif(beta, 2),
      "\np = ", signif(p_value, 2)
    )
  )

library(ggplot2)

# Microniche names epithelial
niche_labels <- c(
  "1" = "Microniche1\nAdipose stromal",
  "2" = "Microniche2\nBasal epithelial",
  "3" = "Microniche3\nLuminal epithelial",
  "4" = "Microniche4\nVascular stromal",
  "5" = "Microniche5\nPerivascular stromal",
  "6" = "Microniche6\nInflammatory stromal"
)

# Microniche names stromal
# niche_labels <- c(
#   "1" = "Microniche1\nSmooth muscle-like stromal",
#   "2" = "Microniche2\nEndothelial stromal",
#   "3" = "Microniche3\nLymphatic endothelial",
#   "4" = "Microniche4\nFibroblast stromal",
#   "5" = "Microniche5\nEpithelial niche",
#   "6" = "Microniche6\nInflammatory niche"
# )

# Same six colours as before
niche_cols <- c(
  "1" = "#D1495B",
  "2" = "#EDAE49",
  "3" = "#4C8DB8",
  "4" = "#72B7A1",
  "5" = "#8E6C88",
  "6" = "#6A994E"
)

ggplot(
  ecm_summary,
  aes(
    x = age,
    y = mean_ecm,
    colour = factor(niche),
    fill = factor(niche)
  )
) +
  geom_point(
    size = 1.8,
    alpha = 0.7
  ) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    linewidth = 0.9,
    alpha = 0.18
  ) +
  
  # regression results
  geom_text(
    data = reg_results2,
    aes(
      x = Inf,
      y = Inf,
      label = reg_label
    ),
    hjust = 1.08,
    vjust = 1.15,
    inherit.aes = FALSE,
    colour = "black",
    size = 3.8
  ) +
  
  facet_wrap(
    ~ niche,
    nrow = 1,
    labeller = as_labeller(niche_labels)
  ) +
  
  scale_colour_manual(values = niche_cols) +
  scale_fill_manual(values = niche_cols) +
  
  labs(
    x = "Age",
    y = "ECM remodelling score"
  ) +
  
  theme_classic() +
  theme(
    # panel border
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      linewidth = 0.6
    ),
    
    # facet title box
    strip.background = element_rect(
      fill = "white",
      colour = "black",
      linewidth = 0.6
    ),
    
    strip.text = element_text(
      face = "bold",
      size = 11,
      lineheight = 0.9
    ),
    
    # axes
    axis.title = element_text(
      face = "bold",
      size = 12
    ),
    
    axis.text = element_text(
      face = "bold",
      size = 10
    ),
    
    # reduce gaps between panels
    panel.spacing = unit(0.15, "lines"),
    
    # no legend
    legend.position = "none"
  )

```

### Find marker genes for each micro-niche using limma - Supplementary Table 2 & 3

```{r}

spe <- readRDS("/dski/nobackup/beileib/GHIST2/prediction2_spe_immune/spe_joint_epicompart.rds")

expr_sparse <- assay(spe, "logcounts")
cl <- factor(spe$clust_M0_lam0.8_kmeans6)

cluster_levels <- levels(cl)

cluster_stats <- list()

for (k in cluster_levels) {
  idx <- cl == k
  out_idx <- !idx
  cluster_stats[[k]] <- data.frame(
    gene = rownames(expr_sparse),
    mean_in = Matrix::rowMeans(expr_sparse[, idx, drop = FALSE]),
    mean_out = Matrix::rowMeans(expr_sparse[, out_idx, drop = FALSE]),
    pct_in = Matrix::rowMeans(expr_sparse[, idx, drop = FALSE] > 0),
    pct_out = Matrix::rowMeans(expr_sparse[, out_idx, drop = FALSE] > 0),
    stringsAsFactors = FALSE
  )
}

expr <- as.matrix(expr_sparse)

### design matrix
design <- model.matrix(~ 0 + cl)
colnames(design) <- paste0("niche", cluster_levels)

fit <- lmFit(expr, design)

n <- as.numeric(table(cl))
names(n) <- paste0("niche", cluster_levels)

contrast_mat <- matrix(
  0,
  nrow = length(cluster_levels),
  ncol = length(cluster_levels),
  dimnames = list(
    paste0("niche", cluster_levels),
    paste0("niche", cluster_levels, "_vs_rest")
  )
)

for (k in cluster_levels) {
  cname <- paste0("niche", k)
  rest <- setdiff(paste0("niche", cluster_levels), cname)
  rest_n <- n[rest]

  contrast_mat[cname, paste0(cname, "_vs_rest")] <- 1
  contrast_mat[rest, paste0(cname, "_vs_rest")] <- -rest_n / sum(rest_n)
}

fit2 <- contrasts.fit(fit, contrast_mat)
fit2 <- eBayes(fit2, trend = TRUE, robust = TRUE)

all_top <- list()
summary_rows <- list()

for (k in cluster_levels) {
  coef_name <- paste0("niche", k, "_vs_rest")

  tt <- topTable(
    fit2,
    coef = coef_name,
    number = Inf,
    sort.by = "P"
  )

  tt$gene <- rownames(tt)
  tt <- tt[, c("gene", setdiff(colnames(tt), "gene"))]

  st <- cluster_stats[[k]]
  tt <- merge(tt, st, by = "gene", all.x = TRUE, sort = FALSE)

  tt$delta_mean <- tt$mean_in - tt$mean_out
  tt$delta_pct <- tt$pct_in - tt$pct_out
  tt$niche <- k

  tt <- tt[order(tt$adj.P.Val, -tt$logFC), ]

  sig <- tt[tt$adj.P.Val < 0.05 & tt$logFC > 0, , drop = FALSE]

  all_top[[k]] <- head(tt, 100)

  summary_rows[[k]] <- data.frame(
    niche = k,
    n_cells = sum(cl == k),
    n_markers_fdr05_logfc_pos = nrow(sig),
    top10_genes = paste(head(sig$gene, 10), collapse = ";"),
    stringsAsFactors = FALSE
  )
}

top_combined <- do.call(rbind, all_top)
summary_df <- do.call(rbind, summary_rows)

```

Please note that the order of micro-niches here is different from the original paper which is because of "set.seed" was different on different machines.